\section{Other OSes}

Most Unix-derived Operating Systems (Linux, BSD flavours, OpenSolaris, etc)
integrate their \ac{ahci} subsystem into a larger disk subsystem with support
for IDE disks, \ac{sata} disks (via \ac{ahci}), CDROM/DVD drives, and also
floppy drives.

This larger disk subsystem often utilizes a general buffer layer which the OS
kernel provides to its subsystems. Furthermore most Unix derivates -- due to
their essentially monolithic nature -- couple the different layers of their
disk subsystems (transport layer, message format and disk commands, e.g.
\ac{ahci}, \ac{sata} and \ac{ata} respectively) using function pointers and
most of them have in-kernel structures that describe commands that are issued
to the disk in a message format and transport agnostic way. This makes those
systems relatively easy to extend by adding a new layer implementations (e.g.
when \ac{ahci} was first implemented a few years ago, it was as simple as
providing a new transport layer implementation for disks attached to \ac{ahci}
controller).

\subsection{FreeBSD}

FreeBSD employs the \ac{cam}\footnote{FreeBSD \acs{scsi} Documentation can be
found under
\url{http://www.freebsd.org/doc/en_US.ISO8859-1/books/arch-handbook/scsi-general.html}}
framework to seperate implementation of the driver for the I/O bus from the
device driver for the attached device. Therefore, FreeBSD's \ac{ahci} driver is
realized as a \ac{sim} handling the I/O operations needed to get an \ac{ata} or
\acs{scsi} command to the device (transparently using the packet interface of
\ac{atapi}). Other aspects of the storage system, such as filesystem or disk
driver do not have to be modified.

\subsection{Linux}

Linux handles access to \ac{ata} devices with libATA\footnote{The libATA
Developer's Guide can be found under
\url{http://www.kernel.org/doc/htmldocs/libata.html}} which unifies interfacing
with \acs{scsi} and \ac{ata} based devices and adapters in a common API. libATA
can translate \acs{scsi} commands to \ac{ata} and vice-versa or simulate a
certain command if there is no translation possible.  Drivers for adapters only
need to implement hooks for basic device operations and communication.
