\chapter*{Introduction}
\epigraph{The Filet-o-Fish contains a battered fish patty made mostly
          from pollock and/or hoki.}{Wikipedia}


Filet-o-Fish, abbreviated \emph{FoF} hereafter, is a tool for the
working language designer. Developed in the context of
Barrelfish\cite{mmcs-barrelfish}, FoF aims at easing the development
of Domain-Specific Languages (DSL) as well as enhancing their
safety. As a side effect of FoF's design, it also becomes easier for
the user of a DSL to understand ``what is going on''.

To achieve this goal, Filet-o-Fish defines a set of
\emph{combinators}. A combinator is a Haskell function manipulating
some Haskell data-types. In this case, our combinators manipulate an
abstraction of the C language constructs, such as integers, floats,
structures, arrays, etc. Altogether, this set of combinators defines
an \emph{embedded language} in Haskell. To avoid the confusion with
the DSLs we are willing to implement, we term this embedded language
the \emph{meta-language}. 

You seems confused now. Listen. The Hamlet compiler is implemented
with Filet-o-Fish. Hamlet is a Domain-Specific Language. In Hamlet's
compiler, we use FoF to \emph{get the job done}, ie. to get the actual
C code out of our capability system description. Hence, the Hamlet
compiler is partly developed in the FoF meta-language. Understood?

However, Filet-o-Fish is much more than a language to get the job
done: being able to compile the meta-language to C is just one
side-effect of our work. By writing a DSL compiler with FoF, you
actually define the \emph{semantics} of the DSL. Whereas the syntax
defines the set of legal expressions of a language, the semantics
assign a meaning to the terms of the language. Note that the C
language does not have any formal semantics. And, no, this is not
normal. This is Evil.

For a DSL, the benefit of having a formal semantics is twofold. First,
the semantics of your DSL is the most precise and accurate description
of the behavior of your domain-specific constructs. An informal,
in-English specification of the DSL might fail to capture some
specific points. The formal semantics is an ultimate documentation,
which doesn't lie. Second, defining a formal semantics is a necessary
step before any compiler correctness proof, be it mechanized or on
paper. Therefore, thanks to FoF, you get a formal, mechanized
semantics of your DSL. And this is for free.

Finally, this document is the literate Haskell code of Filet-o-Fish:
the code described in the following pages is the one that is compiled
by the Haskell compiler. Therefore, this is the most accurate,
up-to-date documentation of Fof's internals.

So much marketing, let us look at the code.

